---
title: "Olink Analysis "
author: "Shuntai Zhou"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_float: true
    theme: sandstone
    highlight: haddock
    code_folding: hide
editor_options: 
  markdown: 
    wrap: sentence

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F )
```

```{r}
library(tidyverse)
library(OlinkAnalyze)
library(readxl)
library(patchwork)

```

## Import data 


```{r}
data <- read_csv("data.csv")
```


```{r, eval = F}
data$PlateID %>% unique()
```

```{r, eval = F}
p44 <- data %>% filter (PlateID == "20201017_SS200044_CARDIO_INF" | PlateID == "20201017_SS200044_NEURO_ONC")
p45 <- data %>% filter (PlateID == "20201017_SS200045_NEURO_ONC" | PlateID == "20201017_SS200045_CARDIO_INF")
p47 <- data %>% filter (PlateID == "20201017_SS200047_CARDIO_INF" | PlateID == "20201017_SS200047_NEURO_ONC")
p48 <- data %>% filter (PlateID == "20201017_SS200048_NEURO_ONC" | PlateID == "20201017_SS200048_NEURO_ONC")
```

```{r, eval = F}
data$SampleID %>% unique() %>% sort()
```

```{r, eval = F}
intersect(p44$SampleID, p45$SampleID)
intersect(p47$SampleID, p48$SampleID)
intersect(p44$SampleID, p47$SampleID)
intersect(p48$SampleID, p45$SampleID)
```

```{r, eval = F}
data %>% filter(Assay == "SMOC2",
                NPX == -0.7636)

data %>% filter(Assay == "SMPD1",
                NPX == -0.394)
```

Join the key table with the olink data table

```{r}
key <- read_csv('PID_key.csv')

joined_data <- data %>% full_join(key) 

joined_data <- joined_data %>% filter(!is.na(Group))

joined_data %>% mutate (
  group = case_when(
    Group == 1 ~ "Group 1", 
    Group == 2 ~ "Group 2", 
    .default = "Group 3&4"
  )
) -> joined_data
```

## PCA plot 

```{r}
joined_data %>% 
  olink_pca_plot(df = .,
                 color_g = "QC_Warning", byPanel = TRUE)  
```


## One way non-parametric poc-hoc analysis for all groups


```{r}
joined_data %>% 
  olink_one_non_parametric(
    variable = "group",
    subject = "PID"
  ) -> all_group_comparison

all_group_comparison %>% filter(
  Threshold == 'Significant'
) %>% select(OlinkID) %>% 
  pull() -> significant_assays

joined_data %>% 
  olink_one_non_parametric_posthoc(
    variable = "group", 
    olinkid_list = significant_assays
  ) -> posthoc_results 

write_csv(posthoc_results, "comparisons/posthoc_results_all_comparisons.csv")

posthoc_results %>% filter(Threshold == 'Significant') -> posthoc_results_significant
```


## Compare group 1 vs group 3&4

```{r}
data1 <- joined_data %>% filter(
  group != "Group 2"
)
```

### non-parametricf comparisons 

```{r}
data1 %>% olink_one_non_parametric(
    variable = "group",
    subject = "PID") -> group1vs34 

write_csv(group1vs34, "comparisons/group1vs34.csv")
```


### pathway enrichment

```{r}

ttest_results1 <- olink_ttest(
  df = data1,
  variable = "group",
  alternative = "two.sided")


gsea_results1 <- olink_pathway_enrichment(data = data1, test_results = ttest_results1)
ora_results1 <- olink_pathway_enrichment(
  data = data1,
  test_results = ttest_results1, method = "ORA")

write_csv(gsea_results1, "comparisons/gsea_results_1vs34.csv")
write_csv(ora_results1, "comparisons/ora_results_1vs34.csv")

```



```{r}
theme_set(theme_bw(base_family = "ArialNarrow") + 
            theme(axis.title.y = element_blank(),
                  axis.text.y = element_text(size = 8)))
number_pathways_to_show = 15
```

Plot GSEA and ORA plots

```{r}

gsea_ora_plot <- function(gsea_results, ora_results) {
  list_size = gsea_results %>% nrow()

  for (i in 1:list_size) {
    gsea_results$Count[i] = gsea_results$core_enrichment[i] %>% str_split_1("/") %>% length()
  }
  
  gsea_results <- gsea_results %>% mutate(
    Description = str_replace_all(Description, "_", " ")
  )
  ora_results$Gene_Ratio <- sapply(ora_results$GeneRatio, function(x) eval(parse(text = x))) 
  
  ora_results <- ora_results %>% mutate(
    Description = str_replace_all(Description, "_", " ")
  )
  
  gsea_results_head <- gsea_results  %>% head(number_pathways_to_show) %>% arrange(desc(NES))

  gsea_results_head %>% 
    ggplot(aes(x = NES, y = Description, color = p.adjust)) + 
    geom_point(aes(size = Count)) + 
    scale_y_discrete(limits = gsea_results_head$Description %>% rev(), 
                     labels = function(x) str_wrap(x, width = 30, whitespace_only = F)) +
    scale_colour_gradient(high = "#833ab4", low = "#fcb045") + 
    scale_size_continuous(range = c(3,6)) +
    labs(title = "Gene Set Enrichment\nAnalysis") -> gsea_plot
  
  ora_results_head <- ora_results  %>% head(number_pathways_to_show) %>% arrange(desc(Gene_Ratio))
  
  ora_results_head %>% 
    ggplot(aes(x = Gene_Ratio, y = Description, color = p.adjust)) + 
    geom_point(aes(size = Count)) + 
    scale_y_discrete(limits = ora_results_head$Description %>% rev(), 
                     labels = function(x) str_wrap(x, width = 30, whitespace_only = F)) +
    scale_colour_gradient(high = "#833ab4", low = "#fcb045") + 
    scale_size_continuous(range = c(3,6))  + 
    labs(x = "Gene Ratio", 
         title = "Over-representation\nAnalysis") -> ora_plot
  
  gsea_plot + ora_plot
  
}

```





```{r, fig.width=10, fig.height=10}
p <- gsea_ora_plot(gsea_results1, ora_results1) + 
  plot_annotation(title = "GSEA and ORA analysis between Group 1 and Group 3&4", 
                    subtitle = paste(number_pathways_to_show, "Pathways with lowest p.adjust values"))

p 

ggsave("figures/pathway_1vs34.png", p, width = 12, height = 8)
```


## Compare group 2 vs group 3&4

```{r}
data2 <- joined_data %>% filter(
  group != "Group 1"
)
```

### non-parametricf comparisons 

```{r}
data2 %>% olink_one_non_parametric(
    variable = "group",
    subject = "PID") -> group2vs34 

write_csv(group2vs34, "comparisons/group2vs34.csv")
```

### pathway enrichment

```{r}

ttest_results2 <- olink_ttest(
  df = data2,
  variable = "group",
  alternative = "two.sided")


gsea_results2 <- olink_pathway_enrichment(data = data2, test_results = ttest_results2)
ora_results2 <- olink_pathway_enrichment(
  data = data2,
  test_results = ttest_results2, method = "ORA")

write_csv(gsea_results2, "comparisons/gsea_results_2vs34.csv")
write_csv(ora_results2, "comparisons/ora_results_2vs34.csv")

```
Plot GSEA and ORA plots

```{r, fig.width=10, fig.height=10}
p <- gsea_ora_plot(gsea_results2, ora_results2) + 
  plot_annotation(title = "GSEA and ORA analysis between Group 2 and Group 3&4", 
                    subtitle = paste(number_pathways_to_show, "Pathways with lowest p.adjust values"))

p 

ggsave("figures/pathway_2vs34.png", p, width = 12, height = 8)
```
### Vocalno plots for Group 1 vs Group 3&4

```{r, fig.width= 8, fig.height= 8}
top_10_name <- ttest_results1 %>%
  slice_head(n = 10) %>%
  pull(OlinkID)

olink_volcano_plot(p.val_tbl = ttest_results1,
                   x_lab = 'Differences between Group 1 and Group 3&4\nGroup 1 minus Group 3&4',
                   olinkid_list = top_10_name)

ggsave("figures/vocalno_1vs34.tiff", width = 8, height = 8)
```

### bar plots for up-regulated and down-regulated makers for Group 1 vs Group 3&4

```{r}

updown_bar <- function(ttest_results) {
  top10_upreg <- ttest_results %>% 
    filter(estimate > 0) %>% slice_head(n = 10)
  
  top10_dnreg <- ttest_results %>% 
    filter(estimate < 0) %>% slice_head(n = 10)
  
  up_and_down <- rbind(top10_upreg, top10_dnreg)
  
  up_and_down <- up_and_down %>% arrange(desc(estimate))
  
  up_and_down %>% ggplot(aes(x = estimate, y = Assay, fill = Adjusted_pval)) + 
    geom_col() + 
    scale_y_discrete(limits = rev(up_and_down$Assay)) + 
    scale_fill_gradient(name = "adjusted\np value", high = "#833ab4", low = "#fcb045")
}

p <- updown_bar(ttest_results1) + 
  labs(title = "Top 10 up-regulated and down-regulated markers\nby adjusted p values", 
       subtitle = "Group 1 vs. Group 3&4\n(estimates from group 1 minus estimates from group 3&4)")

p

ggsave("figures/updown_1vs34.tiff", width = 5, height = 5)
```

### Vocalno plots for Group 2 vs Group 3&4

```{r, fig.width= 8, fig.height= 8}
top_10_name <- ttest_results2 %>%
  slice_head(n = 10) %>%
  pull(OlinkID)

olink_volcano_plot(p.val_tbl = ttest_results2,
                   x_lab = 'Differences between Group 2 and Group 3&4\nGroup 2 minus Group 3&4',
                   olinkid_list = top_10_name)

ggsave("figures/vocalno_2vs34.tiff", width = 8, height = 8)
```


### bar plots for up-regulated and down-regulated makers for Group 2 vs Group 3&4

```{r}


p <- updown_bar(ttest_results2) + 
  labs(title = "Top 10 up-regulated and down-regulated markers\nby adjusted p values", 
       subtitle = "Group 2 vs. Group 3&4\n(estimates from group 2 minus estimates from group 3&4)")

p

ggsave("figures/updown_2vs34.tiff", width = 5, height = 5)
```


## Comparison between group 2 and group 1


### Pathway enrichment
```{r}

data3 <- joined_data %>% filter(
  group != "Group 3&4"
)

data3 <- data3 %>% mutate(
  group = group %>% factor(levels = c("Group 2", "Group 1"))
)

ttest_results3 <- olink_ttest(
  df = data3,
  variable = "group",
  alternative = "two.sided")


gsea_results3 <- olink_pathway_enrichment(data = data3, test_results = ttest_results3)
#ora_results3 <- olink_pathway_enrichment(
  #data = data3,
  #test_results = ttest_results3, method = "ORA")

write_csv(gsea_results3, "comparisons/gsea_results_2vs1.csv")
#write_csv(ora_results3, "comparisons/ora_results_1vs2.csv")

list_size = gsea_results3 %>% nrow()

for (i in 1:list_size) {
  gsea_results3$Count[i] = gsea_results3$core_enrichment[i] %>% str_split_1("/") %>% length()
}

gsea_results3 <- gsea_results3 %>% mutate(
  Description = str_replace_all(Description, "_", " ")
)


gsea_results_head <- gsea_results3  %>% 
  filter(p.adjust <= 0.05) %>% 
  head(number_pathways_to_show) %>% arrange(desc(NES))

gsea_results_head %>% 
  ggplot(aes(x = NES, y = Description, color = p.adjust)) + 
  geom_point(aes(size = Count)) + 
  scale_y_discrete(limits = gsea_results_head$Description %>% rev(), 
                   labels = function(x) str_wrap(x, width = 40, whitespace_only = F)) +
  scale_colour_gradient(high = "#833ab4", low = "#fcb045") + 
  scale_size_continuous(range = c(1,4)) +
  labs(title = "Gene Set Enrichment Analysis", 
       subtitle = "Group 2 vs. Group 1") -> gsea_plot


gsea_plot

ggsave("figures/gsea_plot_2vs1.png", height = 6, width = 5)
```

```{r}
olink_volcano_plot <- function (p.val_tbl, x_lab = "Estimate", olinkid_list = NULL, ...)
{

  #checking ellipsis
  if(length(list(...)) > 0){

    ellipsis_variables <- names(list(...))

    if(length(ellipsis_variables) == 1){

      if(!(ellipsis_variables == 'coloroption')){

        stop(paste0('The ... option only takes the coloroption argument. ... currently contains the variable ',
                    ellipsis_variables,
                    '.'))

      }

    }else{

      stop(paste0('The ... option only takes one argument. ... currently contains the variables ',
                  paste(ellipsis_variables, collapse = ', '),
                  '.'))
    }
  }

  if(is.null(olinkid_list)){

    olinkid_list <- p.val_tbl %>%
      dplyr::filter(Threshold == 'Significant') %>%
      dplyr::pull(OlinkID)

  }


  volcano_plot <- p.val_tbl %>%
    ggplot2::ggplot(ggplot2::aes(x = estimate, y = -log10(Adjusted_pval),
                                 color = Threshold)) +
    ggplot2::geom_point() +
    ggplot2::labs(x = x_lab, y = "-log10(adjusted p-value)") +
    ggrepel::geom_label_repel(data = subset(p.val_tbl, OlinkID %in% olinkid_list),
                              ggplot2::aes(label = Assay), box.padding = 1, 
                              color = "black", show.legend = FALSE) +
    ggplot2::geom_hline(yintercept = -log10(0.05), linetype="dotted") +
    OlinkAnalyze::set_plot_theme() +
    OlinkAnalyze::olink_color_discrete(...)  + 
    theme(axis.title.x = element_text(size = 20),
          axis.text.x = element_text(size = 14),
          axis.title.y = element_text(size = 20), 
          axis.text.y = element_text(size = 14),)


  return(volcano_plot)

}
```



### Volcano plot for group 2 vs group 1

```{r, fig.width= 8, fig.height= 8}
top_10_name <- ttest_results3 %>%
  slice_head(n = 10) %>%
  pull(OlinkID)

olink_volcano_plot(p.val_tbl = ttest_results3,
                   x_lab = 'Differences between Group 2 and Group 1\nGroup 2 minus Group 1',
                   olinkid_list = top_10_name) 
 
ggsave("figures/vocalno_2vs1.pdf", width = 8, height = 8)
```


## compare group 1&2 vs group 3&4

```{r}
data4 <- joined_data %>% mutate(
  group = ifelse (group == "Group 3&4", group, "Group 1&2"), 
  group = group %>% factor()
)

ttest_results4 <- olink_ttest(
  df = data4,
  variable = "group",
  alternative = "two.sided")
```

```{r}
top_10_name <- ttest_results4 %>%
  slice_head(n = 10) %>%
  pull(OlinkID)

olink_volcano_plot(p.val_tbl = ttest_results4,
                   x_lab = 'Differences between Group 1&2 and Group 3&4\nGroup 1&2 minus Group 3&4',
                   olinkid_list = top_10_name)

ggsave("figures/vocalno_12vs34.pdf", width = 8, height = 8)
```

```{r}

p <- updown_bar(ttest_results4) + 
  labs(title = "Top 10 up-regulated and down-regulated markers\nby adjusted p values", 
       subtitle = "Group 1&2 vs. Group 3&4\n(estimates from group 1&2 minus estimates from group 3&4)")

p

ggsave("figures/updown_12vs34.svg", width = 5, height = 5)




```
### group 12 vs 34, gsea and ora

```{r}
gsea_results4 <- olink_pathway_enrichment(data = data4, test_results = ttest_results4)
ora_results4 <- olink_pathway_enrichment(
  data = data4,
  test_results = ttest_results4, method = "ORA")


```
                                                                                                                                                                                             
```{r}
number_pathways_to_show = 8
p <- gsea_ora_plot(gsea_results4, ora_results4) + 
  plot_annotation(title = "GSEA and ORA analysis between Group 1&2 and Group 3&4", 
                    subtitle = paste(number_pathways_to_show, "Pathways with lowest p.adjust values"))

p 

ggsave("figures/pathway_12vs34.png", p, width = 12, height = 8)
```
                                                                                                                                                                                             
                                                                                                                                                                                             